\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}

\title{A Rank-Count Theory for Local Solutions in the Discretizable Distance Geometry Problem}
\author{Distance Geometry Research Group}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We develop a graph-combinatorial rank-count theory for generic local solution
sets in the Discretizable Distance Geometry Problem. The method replaces
enumeration of Branch-and-Prune branch codes by a rank computation over
\(\mathbb F_2\). For a local active edge set \(F\), graph-derived generators
produce branch masks, and labeled violations record active-edge incompatibilities
with their mirror cliques. The feasible local branch codes form an affine
coset \(s^\ast\oplus\mathcal K_F\), where
\(\mathcal K_F=M(\ker V)\). Consequently, the generic local solution count is
\[
|\Xi_F|
=
2^{
\operatorname{rank}\begin{bmatrix}M\\ V\end{bmatrix}
-
\operatorname{rank}(V)
}.
\]
\end{abstract}

\section{Introduction}

The Discretizable Distance Geometry Problem (DDGP) is a structured form of the
Distance Geometry Problem in \(\mathbb R^K\) in which vertices are placed in a
prescribed order and each non-seed vertex is determined by distances to a set
of \(K\) predecessors \cite{mucherino2012discretizable}.  In the molecular
subclass DMDGP, these predecessors are consecutive, and the resulting
Branch-and-Prune tree has strong partial-reflection symmetries
\cite{lavor2012discretizable,liberti2008branch,liberti2014number,
mucherino2012exploiting,goncalves2021new}.  Those symmetries underlie
generic power-of-two solution counts \cite{liberti2014number},
symmetry-driven enumeration \cite{mucherino2012exploiting}, optimal symmetry
detection \cite{lavor2021optimality}, and symmetry-based build-up algorithms
\cite{goncalves2021new}.

For general DDGP instances, predecessor sets need not be consecutive.
Reflections therefore no longer act on intervals or suffixes of the vertex
order; they propagate through a dependency graph.  This distinction is also
central to the known theory of discretization orders
\cite{cassioli2015discretization,abud2024impossible}.  Moreover, in the
general DDGP, existing negative results show that the number of solutions
cannot be counted from the graph alone in full generality
\cite{abud2024impossible}.  The aim of this paper is therefore more local: we
recover a comparable generic counting theory for fixed local DDGP active
subproblems.  The central replacement is to use graph-derived branch shifts
and to test active-edge compatibility by labeled violation cancellation over
\(\mathbb F_2\).

The contribution is threefold.  First, we define local active subproblems
around a pruning set and represent their branch transformations by cone
generators and base generators.  Second, we introduce a labeled violation
matrix that separates violations caused by different mirror cliques.  Third,
we prove that, under a generic mirror-separation assumption, the feasible local
branch codes form one affine coset whose dimension is computed by a
rank-nullity formula.

All vector spaces, sums, kernels, and ranks below are over \(\mathbb F_2\).
Binary branch-code representations are standard for paths in the
Branch-and-Prune tree; related DDGP/DMDGP binary representations have also been
used to analyze symmetry and degrees of freedom
\cite{mucherino2012exploiting,mucherino2020analysis}.

\section{Local DDGP Subproblems}

Let
\[
V_0=\{v_1,\ldots,v_{K+1}\}
\]
be the fixed initial simplex.  For every non-seed vertex \(i>K+1\), the
predecessor set \(U_i\) has size \(K\).  The predecessor relation defines a
directed acyclic graph with an arc \(u\to i\) whenever \(u\in U_i\).
Related order-dependent reference structures have been studied through
\(K\)-discretization and \(K\)-incident graphs \cite{abud2018k}; here we fix
one local predecessor system and derive branch shifts from its dependency
cones.

\begin{definition}[Fixing set]
The fixing set of a vertex \(i\) is
\[
\operatorname{Fix}_U(i)
=
\{i\}\cup \bigcup_{u\in U_i}\operatorname{Fix}_U(u).
\]
\end{definition}

\begin{definition}[Dependency cone]
The dependency cone of a vertex \(i\) is
\[
\operatorname{Cone}_U(i)
=
\{i\}\cup \bigcup_{w:\,i\in U_w}\operatorname{Cone}_U(w).
\]
\end{definition}

For a set of pruning edges \(P\), define the local vertex set
\[
L_P
=
\bigcup_{\{u,v\}\in P}
\left(\operatorname{Fix}_U(u)\cup \operatorname{Fix}_U(v)\right),
\]
and the local binary decision set
\[
B_P=L_P\setminus V_0.
\]
The set \(B_P\) is the set of branch decisions that can affect the pruning
constraints in \(P\).  Restricting to \(L_P\) keeps the construction local:
vertices outside \(L_P\) do not enter the branch code being counted.

The active edge set is denoted by \(F\).  In the main theorem, \(F\) must
contain the local base edges \(E_0[L_P]\).  In applications one usually takes
\[
F=E_0[L_P]\cup P.
\]
The local solution code is
\[
\Xi_F
=
\{s\in \mathbb F_2^{B_P}: s \text{ realizes every edge in }F\}.
\]

\section{Graph-Derived Generators}

A generator represents a branch transformation together with the mirror clique
that explains its geometric action.  Each generator is a pair
\[
g=(m_g,C_g),
\]
where \(m_g\in\mathbb F_2^{B_P}\) is a branch mask and \(C_g\) is a predecessor
clique.  The mask tells which local branch bits are toggled; the clique tells
which affine mirror is used.

Let \(\mathcal G\) be the generator family and let
\[
A=\mathbb F_2^{\mathcal G}
\]
be the generator coefficient space.

\subsection{Cone Generators}

For each decision vertex \(q\in B_P\), the cone generator is
\[
g_q=(m_q,U_q),
\qquad
\operatorname{supp}(m_q)=B_P\cap \operatorname{Cone}_U(q).
\]
Ordered by the DDGP order, the cone masks form a triangular system with ones on
the diagonal, and therefore span the full local branch space
\(\mathbb F_2^{B_P}\).

\subsection{Base Generators}

For a predecessor clique \(C\), define
\[
\operatorname{Gen}_P(C)=\{w\in B_P: U_w=C\}.
\]
The base generator associated with \(C\) is
\[
g_C=(m_C,C),
\]
where
\[
\operatorname{supp}(m_C)
=
\bigcup_{w\in \operatorname{Gen}_P(C)}
\operatorname{supp}(m_w).
\]
Base generators are useful because two transformations with the same branch
mask may have different geometric meaning if they use different mirror cliques.
Thus a base generator may be redundant as a column of the mask matrix while
still being necessary as a labeled geometric operation.

\subsection{Mask Matrix}

Let \(\mathcal G=\{g^1,\ldots,g^m\}\).  The generator mask matrix is the linear
map
\[
M:A\to \mathbb F_2^{B_P},
\]
whose \(j\)-th column is \(m_{g^j}\).  Thus
\[
M\alpha=\bigoplus_{g^j:\,\alpha_j=1}m_{g^j}.
\]

\section{Labeled Violations}

The mask \(m_g\) says which branch bits change, but it does not by itself say
which distances are preserved.  Preservation depends on the mirror clique
attached to the transformation.

Fix the active edge set \(F\).  Let \(g=(m_g,C_g)\), and write
\[
S_g=\operatorname{supp}(m_g).
\]
An active edge \(e=\{a,b\}\in F\) crosses the support boundary of \(g\) if
\[
|e\cap S_g|=1.
\]
If this happens, one endpoint is moved by the reflection and the other is
fixed.  The edge length is preserved only when the fixed endpoint lies on the
reflection mirror, which is ensured when it belongs to \(C_g\).  Therefore
\(g\) violates \(e\) if
\[
|e\cap S_g|=1
\quad\text{and}\quad
e\setminus S_g\not\subseteq C_g.
\]

Violations are labeled by both the edge and the mirror clique.  Let
\[
\mathcal L
=
\{(e,C_g): g\in\mathcal G,\ e\in F,\ g\text{ violates }e\}.
\]
The labeled violation matrix is the linear map
\[
V:A\to \mathbb F_2^{\mathcal L}.
\]
Rows are indexed by labeled pairs \(\ell_i=(e_i,C_i)\in\mathcal L\), and
columns are indexed by generators \(g^j\).  Its entries are
\[
V_{ij}=1
\quad\Longleftrightarrow\quad
C_i=C_{g^j}\text{ and }g^j\text{ violates }e_i.
\]
Thus \(V\alpha\) is the net labeled violation pattern of the generator
combination \(\alpha\).  The same active edge may be crossed by transformations
using different mirrors; such violations are kept in different rows and cannot
cancel unless their mirror labels agree.

\section{Feasible Branch Shifts}

The branch transformations that preserve all active edges are represented by
generator combinations with zero net labeled violation.  The feasible branch
shift space is
\[
\mathcal K_F=M(\ker V).
\]
Thus \(\mathcal K_F\) consists of branch shifts that can be produced by
zero-violation generator combinations.

The distinction between presentations and branch masks is essential here.
Elements of \(\ker V\) are
generator presentations, not branch masks.  Different generator combinations
may produce the same branch mask.  If \(M\alpha=M\beta\), then
\(\alpha+\beta\in\ker M\).  The object to count is therefore the image
\(M(\ker V)\), not the set of presentations \(\ker V\).

\begin{definition}[Generic mirror separation]
The local active subproblem satisfies generic mirror separation if every branch
shift outside \(\mathcal K_F\) changes at least one active-edge length outside
a proper algebraic exceptional set.
\end{definition}

\section{Main Results}

The first result is purely linear algebra.  It computes the dimension of the
projected zero-violation space without enumerating either local branch strings
or generator presentations.

\begin{lemma}[Branch-shift rank]
For the feasible branch shift space \(\mathcal K_F=M(\ker V)\),
\[
\dim\mathcal K_F
=
\operatorname{rank}\begin{bmatrix}M\\ V\end{bmatrix}
-
\operatorname{rank}(V).
\]
\end{lemma}

\begin{proof}
Let \(m=|\mathcal G|=\dim A\).  Since
\(\mathcal K_F=M(\ker V)\), restrict \(M\) to \(\ker V\):
\[
M|_{\ker V}:\ker V\to \mathbb F_2^{B_P}.
\]
By rank-nullity applied to this restricted map,
\[
\dim\mathcal K_F
=
\dim\ker V
-
\dim(\ker V\cap\ker M).
\]
Now \(\dim\ker V=m-\operatorname{rank}(V)\), and
\[
\ker V\cap\ker M
=
\ker\begin{bmatrix}M\\ V\end{bmatrix}.
\]
Therefore
\[
\dim(\ker V\cap\ker M)
=
m-\operatorname{rank}\begin{bmatrix}M\\ V\end{bmatrix}.
\]
Substitution gives the formula.
\end{proof}

The next lemma is the geometric direction encoded by the labeled violation
matrix: zero net labeled violation gives a composition of mirror-compatible
blocks, hence preserves every active edge.

\begin{lemma}[Zero-violation preservation]
Let \(\alpha\in\ker V\) and let \(h=M\alpha\).  If
\(s\in\Xi_F\), then \(s\oplus h\in\Xi_F\).
\end{lemma}

\begin{proof}
Group the selected generators in \(\alpha\) by mirror clique.  For a fixed
clique \(C\), let \(S_C\) be the XOR of the supports of the selected
generators with mirror \(C\).  The row \((e,C)\) of \(V\alpha\) is the parity
of selected generators with mirror \(C\) that violate the active edge \(e\).
Since \(V\alpha=0\), this parity is zero for every active edge and every
mirror clique.

Therefore the block \((S_C,C)\) is mirror-compatible with each active edge:
if an active edge crosses \(S_C\), its fixed endpoint lies in \(C\).  Indeed,
generator supports are disjoint from their own mirror clique, so an endpoint in
\(C\) is never moved by a generator with mirror \(C\); if a crossing edge had
fixed endpoint outside \(C\), the selected generators with mirror \(C\) would
contribute odd violation parity to the row \((e,C)\).  Reflection through
\(\operatorname{aff}(C)\) preserves edges with both endpoints moved, edges
with neither endpoint moved, and crossing edges whose fixed endpoint lies on
the mirror.  Applying these block reflections clique by clique preserves every
edge in \(F\), and their net branch mask is \(\bigoplus_C S_C=M\alpha=h\).
Thus \(s\oplus h\) realizes every active edge.
\end{proof}

The theorem states that, generically, the converse also holds: every feasible
solution difference must come from such a zero-violation presentation.

\begin{theorem}[Feasible branch shift theorem]
Assume \(K\ge2\), \(E_0[L_P]\subseteq F\), \(\Xi_F\ne\varnothing\), and generic
mirror separation.  Then, for any reference solution \(s^\ast\in\Xi_F\),
\[
\Xi_F
=
s^\ast\oplus \mathcal K_F
=
s^\ast\oplus M(\ker V).
\]
\end{theorem}

\begin{corollary}[Rank count]
Under the assumptions of the theorem,
\[
|\Xi_F|
=
2^{
\operatorname{rank}\begin{bmatrix}M\\ V\end{bmatrix}
-
\operatorname{rank}(V)
}.
\]
\end{corollary}

\begin{proof}
By the theorem, \(\Xi_F\) is an affine coset of \(\mathcal K_F\), so
\(|\Xi_F|=2^{\dim\mathcal K_F}\).  The branch-shift rank lemma gives the
displayed exponent.
\end{proof}

\section{Proof of the Feasible Branch Shift Theorem}

Fix a reference solution \(s^\ast\in\Xi_F\).  We prove that the feasible
branch differences from \(s^\ast\) are exactly \(\mathcal K_F\).

First let \(h\in\mathcal K_F\).  By definition, \(h=M\alpha\) for some
\(\alpha\in\ker V\).  By zero-violation preservation,
\[
s^\ast\oplus h\in\Xi_F.
\]
Hence
\[
s^\ast\oplus\mathcal K_F\subseteq \Xi_F.
\]

Conversely, let \(s\in\Xi_F\) and set
\[
h=s\oplus s^\ast.
\]
The cone generators span the full local branch space
\(\mathbb F_2^{B_P}\), so \(h\) has at least one generator presentation.  If
\(h\notin\mathcal K_F\), then by generic mirror separation this branch shift
changes at least one active-edge length outside a proper algebraic exceptional
set.  This contradicts the fact that both \(s^\ast\) and \(s\) realize every
edge in \(F\).  Therefore \(h\in\mathcal K_F\), and so
\[
\Xi_F\subseteq s^\ast\oplus\mathcal K_F.
\]

The two inclusions give
\[
\Xi_F=s^\ast\oplus\mathcal K_F=s^\ast\oplus M(\ker V),
\]
as claimed.

\section{A Small Worked Example}

We include a fully worked six-vertex example to make the construction
checkable by hand.  Although the instance is small, it contains both cone
generators and base generators, a nontrivial labeled violation kernel, and a
presentation ambiguity that must be removed by the projection \(M(\ker V)\).

Let \(K=2\) and let the fixed initial simplex be
\[
V_0=\{v_1,v_2,v_3\}.
\]
Consider the local predecessor sets
\[
U_4=\{v_1,v_3\},
\qquad
U_5=\{v_1,v_3\},
\qquad
U_6=\{v_1,v_4\},
\]
and pruning edge set
\[
P=\{\{v_3,v_6\},\,\{v_4,v_5\}\}.
\]
The local vertex and decision sets are
\[
L_P=\{v_1,v_3,v_4,v_5,v_6\},
\qquad
B_P=\{v_4,v_5,v_6\}.
\]

There are two predecessor cliques:
\[
C_1=\{v_1,v_3\},\qquad C_2=\{v_1,v_4\}.
\]
The first generates \(v_4\) and \(v_5\); the second generates \(v_6\).  Since
\(v_6\) depends on \(v_4\), we have
\[
\operatorname{Cone}_U(v_4)\cap B_P=\{v_4,v_6\}.
\]
With rows ordered as \((v_4,v_5,v_6)\) and columns ordered as
\[
(g_4,g_5,g_6,g_{C_1},g_{C_2}),
\]
the generator supports are
\[
\begin{array}{c|c|c}
\text{generator} & \text{support} & \text{mirror clique}\\
\hline
g_4 & \{v_4,v_6\} & C_1\\
g_5 & \{v_5\} & C_1\\
g_6 & \{v_6\} & C_2\\
g_{C_1} & \{v_4,v_5,v_6\} & C_1\\
g_{C_2} & \{v_6\} & C_2
\end{array}
\]
and hence
\[
M=
\begin{bmatrix}
1&0&0&1&0\\
0&1&0&1&0\\
1&0&1&1&1
\end{bmatrix}.
\]

The active edge set is \(F=E_0[L_P]\cup P\), where the base edges are
\[
\{v_1,v_4\},\{v_3,v_4\},
\{v_1,v_5\},\{v_3,v_5\},
\{v_1,v_6\},\{v_4,v_6\},
\]
and the pruning edges are \(\{v_3,v_6\}\) and \(\{v_4,v_5\}\).  The nonzero
labeled violations are
\[
(\{v_4,v_5\},C_1)
\quad\text{and}\quad
(\{v_3,v_6\},C_2).
\]
They are obtained as follows.
\[
\begin{array}{c|c|c}
\text{generator} & \text{violated active edge} & \text{label}\\
\hline
g_4 & \{v_4,v_5\} & (\{v_4,v_5\},C_1)\\
g_5 & \{v_4,v_5\} & (\{v_4,v_5\},C_1)\\
g_6 & \{v_3,v_6\} & (\{v_3,v_6\},C_2)\\
g_{C_1} & \text{none} & \text{none}\\
g_{C_2} & \{v_3,v_6\} & (\{v_3,v_6\},C_2)
\end{array}
\]
Thus, with rows ordered as the two labels above,
\[
V=
\begin{bmatrix}
1&1&0&0&0\\
0&0&1&0&1
\end{bmatrix}.
\]
Write a generator coefficient vector as
\[
\alpha=(a_4,a_5,a_6,a_{C_1},a_{C_2}).
\]
Then \(V\alpha=0\) is equivalent to
\[
a_4+a_5=0,
\qquad
a_6+a_{C_2}=0.
\]
Hence \(a_4=a_5=t\), \(a_6=a_{C_2}=u\), and \(a_{C_1}=w\) is free.  Applying
the mask matrix gives
\[
M\alpha
=
t(m_4+m_5)+u(m_6+m_{C_2})+w\,m_{C_1}
=
(t+w)
\begin{bmatrix}
1\\1\\1
\end{bmatrix}.
\]
The parameter \(u\) changes the generator presentation but not the branch mask,
because \(m_6=m_{C_2}\).  Therefore
\[
\mathcal K_F
=
M(\ker V)
=
\left\{
\begin{bmatrix}0\\0\\0\end{bmatrix},
\begin{bmatrix}1\\1\\1\end{bmatrix}
\right\}.
\]
The ranks are
\[
\operatorname{rank}\begin{bmatrix}M\\V\end{bmatrix}=3,
\qquad
\operatorname{rank}(V)=2.
\]
Therefore
\[
|\Xi_F|=2^{3-2}=2.
\]
The nontrivial feasible branch shift is the mask \(\{v_4,v_5,v_6\}\), obtained
from \(g_4\oplus g_5\), or equivalently from the base generator \(g_{C_1}\).
The two feasible local solutions are a reference solution and its simultaneous
reflection through the mirror \(\operatorname{aff}(v_1,v_3)\) on the local
vertices \(v_4,v_5,v_6\).

\section{Scope and Genericity}

The example above is intended as a small reproducible check of the construction:
the generator masks, labeled violation matrix, ranks, and final count can all be
inspected directly.

The generic mirror separation assumption excludes accidental geometric
coincidences.  In non-generic configurations, the actual number of solutions can
be larger than the generic rank count.  This happens when an apparently
violating edge is preserved accidentally, for example because a point lies
unexpectedly on a reflection mirror.  Such cases form a proper algebraic
exceptional set; analogous almost-always qualifications appear in existing
DDGP/DMDGP counting results \cite{liberti2014number,abud2024impossible}.

The case \(K=1\) requires separate treatment because the bracket-separation
argument used for \(K\ge2\) degenerates on the line.

\section{Conclusion}

We have described a rank-count method for generic local DDGP solution sets.  The
method replaces interval partial reflections by graph-derived generators and
replaces individual pruning-edge compatibility by labeled violation
cancellation.  The Feasible Branch Shift Theorem identifies the local solution
code from any reference solution as the affine coset
\[
\Xi_F=s^\ast\oplus M(\ker V),
\]
and the Rank-Count Corollary computes its size using only ranks over
\(\mathbb F_2\).  This gives a combinatorial counting method for fixed local
active DDGP subproblems with non-consecutive predecessor structure, under the
stated generic mirror-separation assumption.

\bibliographystyle{plain}
\bibliography{../references/references}

\end{document}
