\section{Graph-Derived Generators}
\label{sec:generators}

This section defines the graph-derived generators that encode branch transformations and their associated mirror cliques. We introduce cone generators to represent the algebraic propagation of branch decisions within their downstream cones. Base generators are defined for each unique predecessor clique in $\mathcal{C}$. Finally, we construct the generator mask matrix to map linear combinations of these generators to their net branch masks.

A generator represents a branch transformation together with the mirror clique that explains its geometric action. Each generator is a pair
\[
g=(m_g, C_g),
\]
where $m_g \in \mathbb{F}_2^{B_{\mathrm{c}}}$ is a branch mask and $C_g$ is a predecessor clique. The mask tells which pruning-constrained branch bits are toggled; the clique tells which affine mirror is used for the reflection.

\begin{definition}[Cone generator]
\label{def:cone-generator}
For each decision vertex $q \in B_{\mathrm{c}}$, the cone generator is
\[
g_q = (m_q, U_q),
\qquad
\operatorname{supp}(m_q) = B_{\mathrm{c}} \cap \operatorname{Cone}(q).
\]
\end{definition}

When the rows and cone-generator columns are ordered by the DDGP vertex order, the cone-mask matrix is triangular with diagonal entries equal to one. Hence the cone masks $\{m_q \mid q\in B_{\mathrm{c}}\}$ form a basis of the constrained branch space $\mathbb{F}_2^{B_{\mathrm{c}}}$.

The predecessor-clique set used for base generators is
\[
\mathcal{C}=\{U_w\mid w\in B_{\mathrm{c}}\}.
\]

\begin{definition}[Base generator]
\label{def:base-generator}
For a predecessor clique $C\in\mathcal{C}$, define the generator roots:
\[
\operatorname{Gen}(C)=\{w \in B_{\mathrm{c}} \mid U_w = C\}.
\]
The base generator associated with $C$ is
\[
g_C=(m_C, C),
\]
where
\[
\operatorname{supp}(m_C)
=
B_{\mathrm{c}} \cap \bigcup_{w:\,U_w=C} \operatorname{Cone}(w)
=
\bigcup_{w\in \operatorname{Gen}(C)} \operatorname{supp}(m_w).
\]
\end{definition}

Base generators are useful because two transformations with the same branch mask may have different geometric meanings if they use different mirror cliques. 
Specifically, two distinct generators $g_1 = (m, C_1)$ and $g_2 = (m, C_2)$ can share the same algebraic branch mask $m \in \mathbb{F}_2^{B_{\mathrm{c}}}$, representing the same coordinate flips. 
However, if $C_1 \neq C_2$, their mirror cliques span distinct affine hyperplanes in $\mathbb{R}^K$. 
Because a branch flip is realized as a spatial reflection across the hyperplane spanned by its predecessor clique, these generators correspond to different isometric transformations. 
This yields distinct spatial coordinates for the vertices in their propagation cones, altering the Euclidean distances for pruning edges. 
Therefore, algebraic redundancy in the branch masks does not imply geometric equivalence, and both generators must be treated as distinct labeled operations.

After defining both kinds of generators, set $\mathcal{G}$ to be the labeled generator family
\[
\mathcal{G}
=
\{g_q\mid q\in B_{\mathrm{c}}\}\sqcup\{g_C\mid C\in\mathcal{C}\},
\qquad
m=\lvert\mathcal{G}\rvert.
\]
This is an indexed family of generator columns, not a quotient by branch mask. Duplicate masks are retained as separate columns when they carry distinct cone or base labels, and in particular when they carry distinct mirror labels. The generator coefficient space is
\[
A = \mathbb{F}_2^{\mathcal{G}}.
\]
For a mirror clique $C$, write
\[
\mathcal{G}_C=\{g\in\mathcal{G}: C_g=C\}.
\]
Choose an enumeration $\mathcal{G}=\{g^1,\ldots,g^m\}$ for the matrix representation below.

\begin{definition}[Mask matrix]
\label{def:mask-matrix}
The generator mask matrix is the linear map
\[
M: A \to \mathbb{F}_2^{B_{\mathrm{c}}},
\]
whose $j$-th column is $m_{g^j}$. For any generator coefficient vector $\alpha=(\alpha_1,\ldots,\alpha_m)\in A$, the action of $M$ is
\[
M\alpha = \bigoplus_{g^j:\,\alpha_j = 1} m_{g^j}.
\]
\end{definition}

The vector $M\alpha$ represents the net branch mask produced by the generator combination $\alpha$.
