\section{Related Work}
\label{sec:related}

We review three key areas of prior work that directly contextualize our contribution: partial-reflection symmetries in discretized distance geometry, complexity and counting barriers, and the contrast between algorithmic search and algebraic representations.

\subsection{Discretization and Partial-Reflection Symmetries}
The DGP has a continuous search space in general, but certain vertex orders turn it into a discrete search problem. In the DMDGP~\cite{lavor2012discretizable}, each vertex $j>K$ is adjacent to its $K$ immediate predecessors, so $U_j = \{j-K, \ldots, j-1\}$. These predecessor sets are contiguous, and the corresponding local cliques make it possible to place $j$ by $K$-lateration. We call the edges from each vertex to its prescribed predecessors the discretization edges $E_D$; all remaining distance constraints are pruning edges $E_P=E\setminus E_D$, used to test and discard incompatible branches.

Geometrically, each lateration step offers two candidate coordinates. These candidates are mirror images across the affine hyperplane spanned by the predecessor coordinates. Thus a branch choice in the binary search tree can be interpreted as a partial reflection: it fixes the vertices already placed and reflects the remaining suffix of the realization~\cite{mucherino2012exploiting}.

This viewpoint gives the DMDGP search tree a strong algebraic structure. For each admissible branching vertex $v$, one associates a generator $g_v$ that applies the corresponding partial reflection to the suffix $\{v,\ldots,n\}$ of a realization. Composing these generators produces a group action on the realizations explored by the binary search tree. Pruning edges restrict which suffix reflections remain valid symmetries, yielding a subgroup generated by the reflections that do not cross a pruning span~\cite{mucherino2012exploiting,liberti2013counting}.

This symmetry group explains the generic power-of-two solution counts in the exact-distance setting~\cite{liberti2011number,liberti2014number}. Because DMDGP reflections act on suffix intervals of the vertex order, solution counting and search can be described from graph topology rather than numerical edge weights~\cite{lavor2021optimality}. The same symmetries also support build-up, splitting, and search algorithms for DMDGP instances~\cite{mucherino2012exploiting,goncalves2021new}. 

\subsection{Complexity and Counting Barriers}
Relaxing the contiguity requirement of the vertex order leads to important computational trade-offs. Valid contiguous DMDGP orders are NP-complete to find for any fixed dimension $K \ge 2$~\cite{cassioli2015discretization}. In the general DDGP, the predecessor sets $U_j$ need not be consecutive; this broader order structure has been studied through the order-independent $K$-discretization graph and the order-dependent $K$-incident graph~\cite{abud2018k}. Valid DDGP orders, by contrast, can be found by fixed-parameter tractable algorithms with respect to $K$, and in polynomial time for fixed $K$~\cite{abud2024impossible}. However, this relaxation significantly complicates solution counting. 

First, Abud et~al.~\cite{abud2024impossible} proved that no purely combinatorial (weight-independent) counting method exists for the general DDGP: without predecessor cliques, the number of lateration outcomes can depend on the numerical edge weights. The Combinatorial DDGP subclass avoids this barrier by requiring predecessor sets to induce cliques, thereby recovering the positive weight-independent result where the count is almost always $2^{n-K}$ when no additional constraints are present~\cite{abud2024impossible}. Second, when additional constraints are present, counting solutions in the Combinatorial DDGP becomes highly non-trivial. Because predecessors are non-consecutive, branch reflections no longer act on suffix intervals; instead, a branch change at a vertex $q$ propagates dynamically through a directed successor dependency graph. This dynamic propagation is the key obstruction that has prevented topological counting in the presence of additional constraints, which we address in this paper.

\subsection{Algorithmic Search versus Algebraic Representations}
Traditional approaches to handling additional constraints (pruning edges) in discretized distance geometry are predominantly algorithmic. The Branch-and-Prune (BP) algorithm~\cite{liberti2008branch,mucherino2012discretizable} performs a depth-first traversal of the binary search tree, placing each new vertex by lateration and pruning a branch as soon as a distance constraint is violated. Symmetry-based variants reduce this traversal by using partial reflections to update candidate realizations across subproblems induced by pruning edges~\cite{goncalves2021new}. While these algorithms are highly effective and exhibit fixed-parameter tractability under certain branching-frequency conditions, they rely on numerical coordinate calculations and tree traversal. 

In contrast, symbolic and algebraic representations have been developed to study the underlying structure of the search space. Binary branch-code representations are standard for modeling paths in the binary search tree, and related binary formulations have been used to analyze symmetries, degrees of freedom, and coordinate-free distance constraints~\cite{mucherino2012exploiting,goncalves2021new,mucherino2020analysis}. Our work builds on this algebraic perspective. Instead of performing tree-based coordinate searches, we model the feasible branch decisions as an affine space over the finite field $\mathbb{F}_2$. By encoding branch transformations as graph-derived generators and recording their conflicts with additional constraints in a labeled violation matrix, we show that solution counting under additional constraints can be solved algebraically in a single step using a rank formula.
