\section{Main Results}
\label{sec:results}

The central objective of this section is to establish a precise equivalence between the physical geometry of distance-preserving branch shifts and the algebraic structure of the binary subspace $\mathcal{K}_F \subseteq \mathbb{F}_2^{B_{\mathrm{c}}}$. Specifically, we show that the algebraic space $\mathcal{K}_F$ is constructively feasible—meaning every shift in $\mathcal{K}_F$ preserves all active edge lengths—and that, under a natural generic separation condition, no other distance-preserving shifts can exist.

Geometrically, moving from a branch code $s$ to $s \oplus h$ via a branch shift $h \in \mathbb{F}_2^{B_{\mathrm{c}}}$ corresponds to a sequence of partial reflections; whenever such a constrained shift is added to a full branch code, it is extended by zero on $B_{\mathrm{f}}$. A shift preserves the validity of all active constraints if it is produced by a generator combination with zero net labeled violations, meaning $\alpha \in \ker V$. Because multiple generator combinations can yield the same branch mask, the candidate feasible branch-shift space is the image of this kernel under the mask matrix $M$:
\[
\mathcal{K}_F = M(\ker V).
\]

\begin{lemma}[Branch-shift rank]
\label{lem:rank}
For the feasible branch shift space $\mathcal{K}_F = M(\ker V)$,
\[
\dim \mathcal{K}_F
=
\operatorname{rank}_{\mathbb{F}_2}\begin{bmatrix} M \\ V \end{bmatrix}
-
\operatorname{rank}_{\mathbb{F}_2}(V).
\]
\end{lemma}

\begin{proof}
Let $m = \lvert \mathcal{G} \rvert = \dim A$. Since $\mathcal{K}_F = M(\ker V)$, we restrict $M$ to $\ker V$:
\[
M\big|_{\ker V} : \ker V \to \mathbb{F}_2^{B_{\mathrm{c}}}.
\]
By rank-nullity applied to this restricted map,
\[
\dim \mathcal{K}_F
=
\dim \ker V
-
\dim(\ker V \cap \ker M).
\]
Now $\dim \ker V = m - \operatorname{rank}_{\mathbb{F}_2}(V)$, and
\[
\ker V \cap \ker M
=
\ker\begin{bmatrix} M \\ V \end{bmatrix}.
\]
Therefore,
\[
\dim(\ker V \cap \ker M)
=
m - \operatorname{rank}_{\mathbb{F}_2}\begin{bmatrix} M \\ V \end{bmatrix}.
\]
Substituting these dimensions into the rank-nullity equation gives:
\[
\begin{aligned}
\dim \mathcal{K}_F &= (m - \operatorname{rank}_{\mathbb{F}_2}(V)) - \left(m - \operatorname{rank}_{\mathbb{F}_2}\begin{bmatrix} M \\ V \end{bmatrix}\right) \\
&= \operatorname{rank}_{\mathbb{F}_2}\begin{bmatrix} M \\ V \end{bmatrix} - \operatorname{rank}_{\mathbb{F}_2}(V).
\end{aligned}
\]
This completes the proof.
\end{proof}

The next lemma isolates the local geometric step needed below, showing that algebraic branch flips correspond to valid Euclidean reflections under suitable compatibility conditions.

\begin{definition}[Mirror-compatibility]
\label{def:mirror-compatible}
Let $C$ be a predecessor clique, $S \subseteq B_{\mathrm{c}}$ be a subset of constrained branches, and $F$ be the active edge set. The block $(S,C)$ is \emph{mirror-compatible} with $F$ if, for every active edge $e = \{a,b\} \in F$ that crosses the block boundary (i.e., $\lvert e \cap S \rvert = 1$), the fixed endpoint belongs to $C$, meaning $e \setminus S \subseteq C$.
\end{definition}

Mirror-compatibility is a purely topological and combinatorial requirement on how the block boundary interacts with the active constraints. The next lemma shows that when a block satisfies this combinatorial condition alongside appropriate causality constraints, the algebraic branch operation is geometrically realized by a valid Euclidean reflection that preserves all active distance constraints.

\begin{lemma}[Block admissibility]
\label{lem:block-admissibility}
Let $C$ be a predecessor clique with $r(C)=\min\{q\in B_{\mathrm{c}}: U_q=C\}$, and let $S\subseteq B_{\mathrm{c}}$ satisfy $S\subseteq \{r(C),\ldots,n\}$ and $S\cap C=\varnothing.$ If $(S,C)$ is mirror-compatible with the active edge set $F$ (Definition~\ref{def:mirror-compatible}), then for any nondegenerate branch embedding $x(s)$ (with $s_{\mathrm{c}}=s|_{B_{\mathrm{c}}}$), let $R_C$ be the Euclidean reflection across the affine hull $\operatorname{aff}\{x_c(s) : c \in C\}$. Then the coordinates
\[
y_i =
\begin{cases}
R_C(x_i(s)), & i\in S,\\
x_i(s), & i\notin S
\end{cases}
\]
are obtained by the constrained branch code $s_{\mathrm{c}}\oplus \mathbf{1}_S$, and every active edge length is preserved:
\[
\|y_a-y_b\|_2=\|x_a(s)-x_b(s)\|_2
\qquad
\text{for all } \{a,b\}\in F.
\]
\end{lemma}

\begin{proof}
First consider the discretization edges. If $i\in S$, then $i\in B_{\mathrm{c}}$, hence $i\in L$. Since $L$ is closed under fixing sets, every predecessor $u\in U_i$ also lies in $L$, so the discretization edge $\{u,i\}$ belongs to $E_D[L]\subseteq F$. If $u\notin S$, then $\{u,i\}$ crosses the block boundary, and mirror-compatibility gives $u\in C$. Thus we have $U_i\subseteq S\cup C$. Consequently, the transformed predecessor simplex of $i$ is the image of the original predecessor simplex under the same reflection $R_C$: predecessors in $S$ are reflected, and predecessors in $C$ are fixed by $R_C$.

If $i\notin S$ and $i\notin C$, then no predecessor of $i$ can lie in $S$. Indeed, if some $u\in U_i\cap S$, then the active edge $\{u,i\}\in E_D[L]$ would cross the block boundary, and mirror-compatibility would force $i\in C$, a contradiction. Hence the predecessor simplex of such an $i$ is unchanged.

It follows, by induction in the DDGP order, that the coordinates $y_i$ are produced by valid lateration choices on the active subsystem. Vertices outside $S$ keep the same local lateration choice. Vertices in $S$ have both their coordinate and their predecessor simplex transformed by the reflection $R_C$; since $R_C$ reverses the local orientation across the predecessor affine hull, the corresponding lateration branch bit is toggled. Therefore the resulting constrained branch code is $s_{\mathrm{c}}\oplus \mathbf{1}_S$.

It remains to check active edge lengths. Let $e=\{a,b\}\in F$. If both endpoints belong to $S$, then both are acted on by the same Euclidean isometry $R_C$, so the length of $e$ is preserved. If neither endpoint belongs to $S$, both endpoints are fixed, so the length is preserved. Finally, suppose exactly one endpoint belongs to $S$, say $a\in S$ and $b\notin S$. By mirror-compatibility, $b\in C$, hence $R_C$ fixes $x_b(s)$. Therefore
\[
\|y_a-y_b\|_2
=
\|R_C(x_a(s))-x_b(s)\|_2
=
\|R_C(x_a(s))-R_C(x_b(s))\|_2
=
\|x_a(s)-x_b(s)\|_2.
\]
Thus every edge in $F$ is preserved.
\end{proof}

The following lemma establishes the crucial link between the algebraic kernel of the violation matrix and the geometric compatibility of the corresponding deformed blocks. Specifically, any selection of generators in the kernel of $V$ yields a set of blocks that are combinatorially mirror-compatible with the active constraints.

\begin{lemma}[Kernel-induced mirror-compatibility]
\label{lem:kernel-compatibility}
Let $\alpha \in \ker V$. For any mirror clique $C$, let $\alpha_C = \{g \in \mathcal{G}_C : \alpha_g = 1\}$ be the set of selected generators with mirror $C$, and let $S_C = \bigoplus_{g \in \alpha_C} S_g$ be the symmetric difference of their supports. Then the block $(S_C, C)$ is mirror-compatible with the active edge set $F$ in the sense of Definition~\ref{def:mirror-compatible}.
\end{lemma}

\begin{proof}
Suppose that an active edge $e = \{a,b\} \in F$ crosses $S_C$, meaning that exactly one endpoint lies in $S_C$. Without loss of generality, assume $a \in S_C$ and $b \notin S_C$. Let $\chi_g: V \to \{0,1\}$ denote the indicator function of $S_g$ for each $g \in \alpha_C$. The indicator function of $S_C$ satisfies
\[
\chi_{S_C}(v) \equiv \sum_{g \in \alpha_C} \chi_g(v) \pmod 2.
\]
Because generator supports are disjoint from their own mirror cliques, we have $S_g \cap C = \varnothing$ for all $g \in \alpha_C$, which implies $S_C \cap C = \varnothing$ and hence $a \notin C$.

Now, suppose for contradiction that the fixed endpoint $b$ is also not in $C$. A generator $g \in \alpha_C$ violates $e$ with label $C$ if and only if it crosses $e$ (i.e., $\lvert e \cap S_g \rvert = 1$) and the fixed endpoint $e \setminus S_g$ lies outside $C$. Since $a \notin C$ and $b \notin C$, the latter condition is satisfied for every crossing generator. The number of such crossing generators in $\alpha_C$ is
\[
\sum_{g \in \alpha_C} \lvert e \cap S_g \rvert
=
\sum_{g \in \alpha_C} (\chi_g(a) + \chi_g(b))
\equiv
\chi_{S_C}(a) + \chi_{S_C}(b)
\equiv
1 + 0
\equiv
1 \pmod 2.
\]
Thus, an odd number of generators in $\alpha_C$ violate $e$ with label $C$, contributing an odd parity to the row $(e,C)$ of the violation product $V\alpha$. This implies $(V\alpha)_{(e,C)} \equiv 1 \pmod 2$, which directly contradicts the assumption that $\alpha \in \ker V$ (which requires $V\alpha = 0$). Hence, the fixed endpoint $b$ must lie in $C$, establishing that the block $(S_C, C)$ is mirror-compatible with $F$.
\end{proof}

Using this algebraic-to-geometric bridge, we can now prove that any branch shift in the range of the kernel of $V$ preserves all active edge lengths, meaning it maps valid configurations to valid configurations.

\begin{lemma}[Zero-violation preservation]
\label{lem:preservation}
Let $\alpha \in \ker V$ and let $h = M\alpha$. If $s_{\mathrm{c}} \in \Xi_F$, then $s_{\mathrm{c}} \oplus h \in \Xi_F$.
\end{lemma}

\begin{proof}
Fix a full extension $s\in\mathbb{F}_2^B$ of $s_{\mathrm{c}}$; active-edge lengths are independent of free bits. We group the selected generators in $\alpha$ by mirror clique. For each mirror clique $C$ that appears among the generators, let $\alpha_C = \{g \in \mathcal{G}_C : \alpha_g = 1\}$ and let $S_C = \bigoplus_{g \in \alpha_C} S_g$ be the symmetric difference of their supports. By Lemma~\ref{lem:kernel-compatibility}, the block $(S_C, C)$ is mirror-compatible with the active edge set $F$. Moreover, since $S_g \cap C = \varnothing$ for all $g \in \alpha_C$, we have $S_C \cap C = \varnothing$.

For any mirror clique $C$ that appears among the generators, let
\[
r(C)=\min\{q\in B_{\mathrm{c}}: U_q=C\}
\]
be the first constrained vertex generated from $C$. Order the nonzero mirror cliques $C_1,\ldots,C_k$ so that $r(C_1)<\cdots<r(C_k)$. Because the support of any generator with mirror $C_i$ only contains vertices discretized at or after $r(C_i)$, we have $S_{C_i}\subseteq\{r(C_i),\ldots,n\}$, so Lemma~\ref{lem:block-admissibility} applies to each block.

Apply the corresponding admissible blocks in reverse order $C_k,\ldots,C_1$. If $j>i$, then $S_{C_j}\subseteq\{r(C_j),\ldots,n\}$ while $C_i\subseteq\{1,\ldots,r(C_i)-1\}$, so $S_{C_j}\cap C_i=\varnothing$. Thus later blocks do not move the mirror clique needed by earlier blocks. Each application preserves all active edge lengths by Lemma~\ref{lem:block-admissibility}. The accumulated constrained branch mask is
\[
\bigoplus_C \mathbf{1}_{S_C}
=
\bigoplus_{g:\,\alpha_g=1} m_g
=
\bigoplus_{g^j:\,\alpha_j=1} m_{g^j}
=
M\alpha
=
h.
\]
Therefore the final constrained code is $s_{\mathrm{c}}\oplus h$, and since all active edge lengths have been preserved, $s_{\mathrm{c}}\oplus h\in\Xi_F$.
\end{proof}

Lemma~\ref{lem:preservation} proves the constructive inclusion: shifts in $\mathcal{K}_F$ generate feasible constrained codes from any feasible reference code. The converse could fail only if a shift outside $\mathcal{K}_F$ preserved all active distances by an accidental geometric coincidence. The following hypothesis excludes exactly those extra solutions.

\begin{definition}[Nondegenerate DDGP parameters]
\label{def:nondegenerate-params}
Fix an ordered DDGP template, which determines $M$, $V$, and $\mathcal{K}_F = M(\ker V)$. A parameter choice $\theta$ for this template is nondegenerate if every lateration step is well defined and, whenever the predecessors of a non-seed vertex have been placed, the corresponding intersection of $K$ spheres consists of two distinct reflected choices. 
\end{definition}

We denote by $\Theta \subset \mathbb{R}^M$ the open, dense set of such nondegenerate parameter choices. For any parameter vector $\theta \in \Theta$, we write $x^\theta(s)$ for the unique physical lateration embedding associated with the branch code $s \in \mathbb{F}_2^B$.

While nondegeneracy ensures that the binary search tree of realizations is topologically well-defined at every step, it is a local condition. It does not prevent "accidental" global distance preservation where two distinct branches, separated by a reflection shift $h \notin \mathcal{K}_F$ not justified by the symmetry group, happen to satisfy all pruning constraints due to an extraordinary numerical alignment of the parameters. To mathematically rule out these exceptional coordinate alignments and guarantee that the algebraic kernel $\mathcal{K}_F$ characterizes \emph{exactly} the viable branch moves, we introduce a stronger, global geometric requirement.

\begin{definition}[Mirror-separated parameters]
\label{def:mirror-sep}
A nondegenerate parameter choice $\theta \in \Theta$ is mirror-separated if, for every constrained shift $h\in\mathbb{F}_2^{B_{\mathrm{c}}}\setminus\mathcal{K}_F$ (extended by zero on $B_{\mathrm{f}}$) and every full branch code $s\in\mathbb{F}_2^B$, there exists an active edge $\{a,b\} \in F$ such that
\[
\|x^\theta_a(s \oplus h) - x^\theta_b(s \oplus h)\|_2^2 \neq \|x^\theta_a(s) - x^\theta_b(s)\|_2^2.
\]
\end{definition}

\begin{remark}[A genericity criterion for mirror separation]
\label{rem:mirror-separation-genericity}
Mirror separation rules out accidental geometric coincidences in which a shift $h\notin\mathcal{K}_F$ nevertheless preserves all active edge distances. Fix the ordered template, the active subsystem $L$, and the active edge set $F$. For each constrained branch shift $h\in\mathbb{F}_2^{B_{\mathrm{c}}}\setminus\mathcal{K}_F$ and active edge $e=\{a,b\}\in F$, consider the discrepancy function
\[
\Delta_{h,e}(\theta;s)
=
\|x^\theta_a(s\oplus h)-x^\theta_b(s\oplus h)\|_2^2
-
\|x^\theta_a(s)-x^\theta_b(s)\|_2^2,
\]
defined on the nondegenerate parameter choices $\theta\in\Theta$ for which the corresponding branch embeddings exist. Mirror separation fails exactly when, for some shift $h\notin\mathcal{K}_F$ and some branch code $s$, all active-edge discrepancies vanish:
\[
\Delta_{h,e}(\theta;s)=0
\qquad
\text{for every } e\in F.
\]
Consequently, if for every $h\notin\mathcal{K}_F$ and every branch code $s$ at least one discrepancy $\Delta_{h,e}(\cdot;s)$ is not identically zero on the relevant nondegenerate parameter component, then the failure set for mirror separation is contained in a finite union of proper exceptional sets defined by the vanishing of nonzero discrepancy functions. In particular, mirror separation holds for almost all such nondegenerate parameters, in the same Lebesgue-measure sense used for generic DMDGP statements and zero-measure pruning coincidences in~\cite{liberti2012discretizable}.
\end{remark}

Finally, the main theorem states that, under mirror separation, the converse also holds on the pruning-constrained branch set. The full count then follows by multiplying by the free branch decisions in $B_{\mathrm{f}}$.

\begin{theorem}[Rank-count theorem]
\label{thm:main}
Assume $K \ge 2$ and $\Xi \ne \varnothing$. Assume also that the present predecessor and edge-length parameters are mirror-separated in the sense of Definition~\ref{def:mirror-sep}. Then, for any reference solution $s^\ast \in \Xi$, with constrained restriction $s^\ast_{\mathrm{c}}=s^\ast|_{B_{\mathrm{c}}}$,
\[
\Xi_F
=
s^\ast_{\mathrm{c}} \oplus \mathcal{K}_F
=
s^\ast_{\mathrm{c}} \oplus M(\ker V).
\]
Consequently,
\[
\lvert \Xi\rvert
=
2^{
f+
\operatorname{rank}_{\mathbb{F}_2}\begin{bmatrix} M \\ V \end{bmatrix}
-
\operatorname{rank}_{\mathbb{F}_2}(V)
}.
\]
\end{theorem}

\begin{proof}
Fix $s^\ast \in \Xi$ and let $s^\ast_{\mathrm{c}} = s^\ast|_{B_{\mathrm{c}}}$.

$(\supseteq)$ For any $h \in \mathcal{K}_F$, Lemma~\ref{lem:preservation} gives
$s^\ast_{\mathrm{c}} \oplus h \in \Xi_F$, so $s^\ast_{\mathrm{c}} \oplus \mathcal{K}_F \subseteq \Xi_F$.

$(\subseteq)$ Let $s_{\mathrm{c}} \in \Xi_F$ and set $h = s_{\mathrm{c}} \oplus s^\ast_{\mathrm{c}}$.
Extend $h$ by zero on $B_{\mathrm{f}}$ and put $\widetilde{s}=s^\ast\oplus h$. Then $\widetilde{s}$ has constrained part $s_{\mathrm{c}}$ and is active-feasible. If $h \notin \mathcal{K}_F$, mirror separation gives an active-edge length difference between $s^\ast$ and $\widetilde{s}$, a contradiction. Hence $h \in \mathcal{K}_F$ and
$\Xi_F \subseteq s^\ast_{\mathrm{c}} \oplus \mathcal{K}_F$.

Together, $\Xi_F = s^\ast_{\mathrm{c}} \oplus M(\ker V)$. Since free branches on
$B_{\mathrm{f}}$ combine independently, $\lvert\Xi\rvert = 2^f \lvert\Xi_F\rvert
= 2^{f + \dim\mathcal{K}_F}$, and Lemma~\ref{lem:rank} gives the rank formula.
\end{proof}
