\section{Feasible Branch Shifts}
\label{sec:shifts}

This section formalizes the concept of the feasible branch shift space, which characterizes the set of valid coordinate transformations. A DDGP realization is parameterized by a branch code $s \in \mathbb{F}_2^{B_{\mathrm{c}}}$. Moving from one branch code to another via a branch shift $h \in \mathbb{F}_2^{B_{\mathrm{c}}}$, written as $s \to s \oplus h$, corresponds geometrically to a sequence of partial reflections across predecessor mirror hyperplanes. A central question in discretization theory is which shifts $h$ ``preserve'' the validity of the realization—meaning that if $s$ yields a valid spatial embedding satisfying the active distance constraints in $F$, the shifted code $s \oplus h$ does so as well. To address this, we leverage the algebraic structure of generator combinations $\alpha$ in the presentation space $A = \mathbb{F}_2^{\mathcal{G}}$. The algebraic net labeled violation of such a combination is given by $V\alpha \in \mathbb{F}_2^{\mathcal{L}}$. If a combination has zero net labeled violations, i.e., $\alpha \in \ker V$, then all individual labeled violations produced on active edges by its constituent generators algebraically cancel out over $\mathbb{F}_2$. Because the labeled violations are carefully indexed by both the active edge and the specific mirror clique, this algebraic cancellation over $\mathbb{F}_2$ prevents spurious interactions between geometrically distinct mirrors, ensuring that the physical distance constraints on all active edges are preserved.

However, a crucial distinction must be maintained between the presentation space $A$—the coefficient space where $\ker V$ resides—and the actual coordinate branch space $\mathbb{F}_2^{B_{\mathrm{c}}}$ where the physical branch shifts $h$ live. The connection between these spaces is given by the mask matrix $M: A \to \mathbb{F}_2^{B_{\mathrm{c}}}$, which maps generator combinations to their net branch masks. Because multiple distinct generator combinations can induce the exact same physical branch shift (due to algebraic redundancy, i.e., a non-trivial kernel $\ker M$), the physical branch shifts that preserve the active distance constraints are represented by the image under $M$ of the kernel of $V$. Thus, we define the feasible branch shift space $\mathcal{K}_F$ as
\[
\mathcal{K}_F = M(\ker V).
\]
This algebraic characterization sets the stage for the \emph{generic mirror separation} hypothesis (Definition~\ref{def:mirror-sep}), which serves as the physical bridge of our theory. While any shift in $\mathcal{K}_F$ is algebraically guaranteed to preserve active edge distances, the converse—that only shifts in $\mathcal{K}_F$ preserve these distances—requires ensuring that no accidental or degenerate spatial alignments allow an infeasible shift $h \notin \mathcal{K}_F$ to preserve all constraints. The generic mirror separation hypothesis guarantees that under generic geometric configurations, any shift not in $\mathcal{K}_F$ will physically violate at least one active edge constraint. Thus, it mathematically seals the equivalence between physical distance-preserving transformations and the algebraic space $\mathcal{K}_F$.

\begin{definition}[Generic mirror separation]
\label{def:mirror-sep}
View the predecessor data and the edge lengths in $F$ as parameters on the real algebraic set where all required lateration steps are nondegenerate. The instance satisfies generic mirror separation if there is a proper algebraic exceptional set $Z$ such that, for every parameter choice outside $Z$, every shift $h\in\mathbb{F}_2^{B_{\mathrm{c}}}\setminus\mathcal{K}_F$, and every constrained branch code $s\in\mathbb{F}_2^{B_{\mathrm{c}}}$ for which the lateration embeddings $x(s)$ and $x(s\oplus h)$ are defined, the two embeddings assign different lengths to at least one active edge in $F$.
\end{definition}

We use generic mirror separation as an explicit instance-level hypothesis. The main theorem below does not assert that every ordered instance satisfies it; rather, it proves the rank count for instances that do.

The next section uses this hypothesis to prove the algebraic rank count.
