\section{Branch Codes and Pruning Constraints}
\label{sec:subproblems}

This section sets up the combinatorial objects used in the counting argument. We first fix the ordered predecessor system that produces the branch decisions, then use its dependency structure to identify which decisions can influence pruning constraints. This separates the branch-code space into constrained and free coordinates, and yields a smaller constrained code whose cardinality determines the total number of realizations.

Let the chosen order be built into the vertex labels, so that $V=\{1,\ldots,n\}$. For each vertex $i$, write $U_i$ for the predecessor set of $i$. Additionally, let $V_0=\{1,\ldots,K\}$ be the fixed initial clique in $\mathbb{R}^K$. For $i\le K$, set $U_i=\varnothing$. For every non-seed vertex $i > K$, the predecessor set $U_i \subseteq \{1,\ldots,i-1\}$ has size $K$. Note that the predecessor relation defines a directed acyclic graph with an arc $j \to i$ whenever $j \in U_i$.

For the rest of the paper, the ordered predecessor family $U=(U_i)_{i\in V}$ is fixed. All predecessor closures are taken with respect to this family, so we suppress the subscript $U$ and write $\operatorname{Fix}$ and $\operatorname{Cone}$.

Two closures of this predecessor relation will be used throughout the construction. The first one looks backward: to determine the position of a vertex, we must also determine the predecessors needed to place it, their predecessors, and so on.

\begin{definition}[Fixing set]
\label{def:fixing}
The fixing set of a vertex $i$ is
\[
\operatorname{Fix}(i)
=
\{i\}\cup \bigcup_{u\in U_i}\operatorname{Fix}(u).
\]
For seed vertices this gives the base case $\operatorname{Fix}(i)=\{i\}$ because $U_i=\varnothing$.
\end{definition}

The forward closure captures propagation: changing a branch decision at a vertex can affect every later vertex that depends on it, directly or indirectly.

\begin{definition}[Dependency cone]
\label{def:cone}
The dependency cone of a vertex $i$ is
\[
\operatorname{Cone}(i)
=
\{i\}\cup \bigcup_{w:\,i\in U_w}\operatorname{Cone}(w),
\]
where $w$ ranges over vertices. If no later vertex has $i$ as a predecessor, this gives the terminal base case $\operatorname{Cone}(i)=\{i\}$.
\end{definition}

Let $E_D$ denote the discretization edges induced by the predecessor relation, and let $E_P = E\setminus E_D$ denote the pruning edges. The full branch-decision set is
\[
B = V\setminus V_0.
\]
Define the pruning-relevant vertex set
\[
L
=
\bigcup_{\{u,v\}\in E_P}
\left(\operatorname{Fix}(u)\cup \operatorname{Fix}(v)\right),
\]
and the pruning-constrained branch set
\[
B_{\mathrm{c}}=L\setminus V_0.
\]
The remaining branch decisions
\[
B_{\mathrm{f}}=B\setminus B_{\mathrm{c}},
\qquad
f=\lvert B_{\mathrm{f}}\rvert,
\]
are free with respect to pruning constraints: changing them preserves all discretization edges by construction and cannot change the endpoints of any pruning edge.

The active edge set is
\[
F = E_D[L] \cup E_P,
\qquad
E_D[L]=\{\{u,v\}\in E_D \mid u,v\in L\}.
\]
When $E_P=\varnothing$, we have $L=\varnothing$, $B_{\mathrm{c}}=\varnothing$, and all branch decisions are free.

Let $x(s)=(x_a(s))_{a\in V}$ denote the embedding determined by the full branch code $s\in\mathbb{F}_2^B$. Throughout, counts are branch-code counts relative to the fixed seed clique $V_0$; realizations are not additionally quotient-identified by global Euclidean congruences. The full solution code is
\[
\Xi
=
\{s \in \mathbb{F}_2^B \mid
\|x_a(s)-x_b(s)\|_2=d_{ab}
\text{ for every } \{a,b\}\in E\}.
\]
The constrained solution code is
\[
\Xi_F
=
\{s_{\mathrm{c}} \in \mathbb{F}_2^{B_{\mathrm{c}}} \mid
\|x_a(s)-x_b(s)\|_2=d_{ab}
\text{ for every } \{a,b\}\in F\},
\]
where $s$ is any full branch code whose restriction to $B_{\mathrm{c}}$ is $s_{\mathrm{c}}$. This is well-defined because branch decisions in $B_{\mathrm{f}}$ do not affect edges in $F$. Hence
\[
\lvert \Xi\rvert = 2^f \lvert \Xi_F\rvert.
\]
Our goal is to compute $\lvert \Xi\rvert$ under generic geometric assumptions without running tree-based branch search.
