\section{Labeled Violations}
\label{sec:violations}

In the discretization of the Distance Geometry Problem (DGP), the distance constraints corresponding to active edges are maintained or violated under partial reflection operations (generator moves). Geometrically, a generator $g = (m_g, C_g)$ acts as an isometry (a reflection) on the vertices within its support $S_g = \operatorname{supp}(m_g)$, while leaving vertices outside $S_g$ stationary. For any active edge $e = \{a,b\}$, its length is trivially preserved if both endpoints are fixed ($a,b \notin S_g$), or if both endpoints are reflected ($a,b \in S_g$), since the reflection is a rigid isometric transformation. Consequently, the distance between $a$ and $b$ is subject to change if and only if $e$ crosses the support boundary of $g$, meaning exactly one endpoint is reflected while the other remains fixed.

Even when an active edge crosses the support boundary, say with $a \in S_g$ reflected and $b \notin S_g$ fixed, the distance between them is preserved if the stationary endpoint $b$ lies on the reflection mirror hyperplane $\operatorname{aff}(C_g)$. Under the assumption of generic non-incidence, where no active vertex outside the predecessor clique $C_g$ accidentally lies on its affine span, this geometric containment is equivalent to the combinatorial condition $b \in C_g$. If $b \notin C_g$, the reflection alters the distance, producing a violation. Crucially, to construct a rigorous linear algebraic counting theory over $\mathbb{F}_2$, we cannot treat violations merely as binary events on edges. Instead, each violation must be tracked as a \emph{labeled violation} indexed by the pair $(e, C_g)$. This dual indexing is an algebraic necessity: two different generator moves violating the same active edge $e$ can only cancel each other's violation algebraically if they reflect across the same mirror hyperplane, i.e., they share the same mirror label $C_g$. Without this labeling, cancellations would falsely occur between independent geometric constraints, destroying the correspondence between the kernel of the violation matrix and valid realization branches.

Fix the active edge set $F$. Let $g=(m_g, C_g)$ be a generator, and write
\[
S_g = \operatorname{supp}(m_g).
\]
An active edge $e = \{a,b\} \in F$ crosses the support boundary of $g$ if exactly one endpoint is toggled:
\[
\lvert e \cap S_g \rvert = 1.
\]
If this happens, one endpoint is moved by the reflection while the other is fixed. For the generic parameter choices considered below, no active vertex outside a predecessor clique $C$ lies accidentally in $\operatorname{aff}(C)$. Under this generic non-incidence condition, the edge length is preserved exactly when the fixed endpoint belongs to $C_g$. Therefore, we say that $g$ violates $e$ with mirror label $C_g$ if
\[
\lvert e \cap S_g \rvert = 1
\quad\text{and}\quad
e\setminus S_g \not\subseteq C_g.
\]
Because $e$ crosses the support boundary, the set difference $e \setminus S_g$ is a singleton containing the unique endpoint of $e$ that remains fixed. Under generic non-incidence, the condition $e \setminus S_g \not\subseteq C_g$ is therefore equivalent to saying that this fixed endpoint does not lie on the reflection mirror hyperplane $\operatorname{aff}(C_g)$.

The labeled violations are indexed by both the edge and the mirror clique. Let
\[
\mathcal{L}
=
\{(e, C_g) \mid g \in \mathcal{G},\ e \in F,\ g \text{ violates } e\}.
\]
The labeled violation matrix is the linear map
\[
V: A \to \mathbb{F}_2^{\mathcal{L}}.
\]
Rows of $V$ are indexed by labeled pairs $\ell_i=(e_i,C_i)\in\mathcal{L}$, and columns are indexed by generators $g^j$. The entries of $V$ are
\[
V_{ij} = 1
\quad\Longleftrightarrow\quad
C_i = C_{g^j} \quad\text{and}\quad g^j \text{ violates } e_i.
\]
Thus, $V\alpha$ represents the net labeled violation pattern of the generator combination $\alpha$. The same active edge may be crossed by transformations using different predecessor mirrors; such violations are kept in distinct rows and cannot cancel unless their mirror labels agree.
